Effects of the body force on the pedestrian and the evacuation dynamics

### 2020-02-07 paper.tex
correcciones en conclusion

\title{Effects of the body force on the pedestrian and evacuation dynamics}

@@

The maximum (realistic) compressions may be computed from the hook relation 
$F(x)=k_n(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force  $10\,$N $-$ $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. Thus, according to 
Table~\ref{table_data}, \textcolor{blue}{we may expect experimental values for $k_n(x)$ up to 
$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N).} \\

@@

The value of $k_n$ affects the velocity all along the corridor. Fig.~\ref{vx_profile} shows the velocity profile ($v_x$ vs. the longitudinal coordinate $y$) for different $k_n$ values. For $k_n=0$, we can see a parabolic-like velocity profile that resembles the Poiseuille flow (for Newtonian and incompressible fluids in laminar flow flowing through a pipe). This behavior was observed in empirical measurements reported in Ref.~\cite{zhang2013empirical}. As $k_n$ increases the velocity profile flattens until becoming a constant ($k_n=1.2\,$E6). In this scenario, the value of mean $v_x$ is much lower than on the case of soft pedestrians ($k_n=$0).\\

If the stiffness is high enough, the crowd behaves like a solid; this means that the crowd can not be easily ``deformed''. In this context, to deform means to have certain parts of the crowd that manage to move faster than other parts of the crowd. \\

The strain rate is a tensor for every point in the space. It expresses how the relative velocity of a medium changes when one moves away from a position in a specific direction. For this particular problem, we define the following version of the strain rate:

@@

Where $\langle \rangle$ means the average taken over time. By this definition we compare the velocity of the pedestrians close to the wall (boundary) with respect to the velocity of the pedestrians in the center of the corridor (center). As the stiffness level increases, the crowd becomes less deformable. This phenomenon is shown in Fig.~\ref{strain_rate_vs_kn} where we can see that the strain rate decreases with $k_n$. \\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./vx_profile.eps}\label{vx_profile}}\ 
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./strain_rate_vs_kn.eps}\label{strain_rate_vs_kn}}\\
\caption[width=0.47\columnwidth]{(a) Velocity in the $x$ coordinate as a function of the transversal coordinate $y$ for three different stiffness values $k_n$ (see the label). (b) Strain rate as a function of the stiffness level $k_n$ for three different global density values (see label). The measurements correspond to a corridor of 28$\,$m $\times$ 22$\,$m with periodic boundary conditions and $v_d=1$. The global density was 7 p$\,$m$^{-2}$. Color online only. }
\label{profile_strain}
\end{figure}

To conclude section \ref{corridor}, we remark that increasing the stiffness produces the opposite results in corridors respect to bottlenecks. The walls in the corridors play a critical role that prevents pedestrians from detaching from each other. This effect has consequences such as increasing the connectivity of the contact network and flattening the velocity profile (thus reducing the strain rate). If the stiffness is high enough, the pedestrians can not push to the pedestrian ahead since the body force repulsion becomes very high. This leads to a ``solidification'' of the crowd that makes all the pedestrians walk at almost the same velocity. The pedestrians that are in contact with the wall are the ones who determine the velocity of the whole crowd. \\

@@

In this section, we inspect different values of the dimensionless numbers $\mathcal{K}_c$ and $\mathcal{K}$ and compare the results with the empirical measurements of Ref.~\cite{helbing_2007}. We varied the body force coefficient ($k_n$) and the friction coefficient ($k_t$) to inspect different values of $\mathcal{K}_c$ and $\mathcal{K}$ respectively.\\

The empirical measurements correspond to the fundamental diagram measured at the entrance of the Jamaraat bridge (see the inset in Fig.~\ref{flow_density_no_body_force}). We aim to reproduce the qualitative behavior of these measurements roughly.\\

Figs. \ref{flow_density} show the flow as a function of the global density (fundamental diagram). Fig.~\ref{flow_density_no_body_force} corresponds to $\mathcal{K}_c=$ 0 (zero body force \textit{i.e.} $k_n=0$) while Fig.~\ref{flow_density_body_force} corresponds to $\mathcal{K}_c=$68 (which is the E6value of the original SFM with $k_n=1.2\,$E5 ). Each curve correspond to a different friction value, hence different $k_t$ value (see the caption).\\

According to most of the empirical measurements of the fundamental diagram, the flow decreases for high enough values of density because the crowd becomes jammed. Notice that the original friction (corresponding to $\mathcal{K}=$137) does not produce a reduction of the flow when there is zero body force ($\mathcal{K}_c=$0). Nevertheless, when the body force is included, and we remain the original friction (circles from Fig.~\ref{flow_density_body_force}), the flow reduces in the range of densities from  5 p m$^{-2}$ to 7 p m$^{-2}$; then increases for higher values of density.\\

When the dimensionless number associated with the friction is $\mathcal{K} = $ 685, the friction is five times the original SFM value. In this scenario, the flow reduces for densities higher than 5 p m$^{-2}$ in both cases (with body force and without body force). The main qualitative difference is that including the body force produces a subtle increment in the flow for densities higher than 7 p m$^{-2}$ (orange curve from Fig.~\ref{flow_density_no_body_force}). \\

When the dimensionless number associated to the friction is $\mathcal{K}=$1371  (the friction is ten times the original SFM value), in this case, the flow reduces or remains constant when the density is higher than 5 p m$^{-2}$ for both cases (with $\mathcal{K}_c=$ 0 and $\mathcal{K}_c = $ 68).\\

It worth mentioning that the SFM contemplates the possibility of overlapping between pedestrians (unlike the models stated in Refs.~\cite{kabalan_2017,jebrane_2019}). Despite this hard-to-interpret feature, the SFM can reproduce empirical behaviors of collective pedestrian dynamics.\\

These results suggest that although increasing $\mathcal{K}_c$ produces a reduction in the flow, it is necessary to increase $\mathcal{K}$ (for instance, increasing the friction coefficient $k_t$) to reproduce the reduction in the flow that characterizes the high-density regime. It is still a pending challenge to find the values of the dimensionless numbers that reproduce the qualitative behavior of the fundamental diagram, but this first approach narrows down the search.\\

@@

In this research, we explore the effect on the pedestrians dynamics of the (sometimes neglected) body force in the framework of the SFM. We showed that the stiffness coefficient ($k_n$) has a significant impact on the evacuation dynamics (bottleneck) and also in the dynamics of pedestrians walking along a straight corridor.\\

In the bottleneck geometry, the evacuation time diminishes (pedestrians move faster) as pedestrians become stiffer for all the ranges of the desired velocities explored. This phenomenon occurs because increasing the stiffness level promotes the repulsion between pedestrians pair interactions. As pedestrians detach from each other, the friction force diminishes due to the reduction of overlap between pedestrians and the contacts among them. The general decrease in frictional force reduces the probability of producing a cluster of pedestrians blocking the exit (blocking cluster), which leads to a more efficient evacuation dynamics. \\

In the corridor geometry, the actual walking speed reduces as pedestrians become stiffer. This is the opposite behavior with respect to the obtained in bottleneck geometry. The difference is explained because although increasing stiffness further separates individuals, they can not detach from each other as much as possible due to the lateral walls (these walls are not relevant in the bottleneck geometry). Thus, when the stiffness increases, the overlap between pedestrians is reduced while the number of contacts (degree) increases. Granular materials can behave like fluid or solids, depending on several conditions. If the stiffness level is low, the entire crowd behaves like a viscous fluid (with a parabolic velocity profile). However, if the stiffness level is high enough, the whole crowd behaves like a crystalline solid with a constant velocity profile which only depends on the friction interaction with the walls.\\

Regarding the parameters of the model, we found that it is possible to reduce the original SFM parameters to three dimensionless parameters. The original parameters interrelate to each other through the dimensionless ones. Many different settings can produce the same dimensionless parameters. Thus, no univocal relations can be established between the individualistic parameters and collective dynamics. In order to explore the outcomes from different dimensionless numbers, we compared numerical results with empirical measurements. We found good candidates to reproduce the qualitative behavior of empirical measurements in the range $0 < \mathcal{K}_c < 68$ with $\mathcal{K} = 685$. Nevertheless, this is a first sketch to arrive at a set of dimensionless parameters that may be suitable to reproduce empirical measurements. We encourage further research that seeks to find values of the dimensionless numbers that best reproduce experimental data.\\

>>>>

\title{Effects of the body force on the pedestrian and the evacuation dynamics}

@@

The maximum (realistic) compressions may be computed from the hook relation 
$F(x)=k_n(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force  $10\,$N $-$ $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. \\
%Thus, according to Table~\ref{table_data}, \textcolor{blue}{we may expect experimental values for $k_n(x)$ up to 
%$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N).} \\

@@

Fig.~\ref{vx_profile} shows the velocity profile ($\langle v_x \rangle$ vs. the transversal coordinate $y$) for different $k_n$ values. For $k_n=0$, we can see a parabolic-like velocity profile that resembles the Poiseuille flow (similar to Newtonian and incompressible fluids in a laminar regime). This behavior was also observed in empirical measurements of crowd dynamics reported in Ref.~\cite{zhang2013empirical}. But as $k_n$ increases the velocity profile flattens until becoming (almost) uniform (see Fig.~\ref{vx_profile} for $k_n=1.2\,$E6). In this scenario, $v_x$ attains a much lower level than in the case of soft pedestrians ($k_n=$0).\\

The crowd behaves like a solid for very stiff pedestrians. This means that the crowd can not be easily ``deformed''. In this context, deformation means that parts of the crowd may be released to move faster than other parts of the crowd. \\

The strain rate is the tensor that expresses how the relative velocity of a medium changes when one moves away from one position to another. We consider the following discrete definition of the strain rate:

@@

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./vx_profile.eps}\label{vx_profile}}\ 
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./strain_rate_vs_kn.eps}\label{strain_rate_vs_kn}}\\
\caption[width=0.47\columnwidth]{(a) Mean velocity in the $x$ coordinate as a function of the transversal coordinate $y$ for three different stiffness values $k_n$ (see the label). (b) Strain rate as a function of the stiffness level $k_n$ for three different global density values (see label). Data correspond to a corridor of 28$\,$m $\times$ 22$\,$m with periodic boundary conditions and $v_d=1$. The global density was $\rho=7 p\,$m$^{-2}$. Color online only. }
\label{profile_strain}
\end{figure}

Where $\langle \cdot \rangle$ means the average taken over time. This definition compares the velocity of the pedestrians close to the wall (boundary) with respect to the velocity of the pedestrians at the center of the corridor (center). Thus, $S$ vanishes (no deformation) as the stiffness level increases. This phenomenon is shown in Fig.~\ref{strain_rate_vs_kn} where we can see that the strain rate drops for high values of $k_n$. \\

We conclude this Section by stressing that stiffer pedestrians attain opposite results in corridors with respect to bottlenecks. The walls in the corridors play a critical role that prevents pedestrians from detaching from each other. This effect has can be observed as an increment in the connectivity of the contact network and the flattening of the velocity profile (thus reducing the strain rate). Very high enough stiffness values stuck the pedestrians leading to a ``solidification'' of the crowd. All the pedestrians walk at almost the same velocity at this stage. The pedestrians that are in contact with the walls are the ones that determine the velocity of the whole crowd. We already analyzed this situation in Ref~\cite{dorso_2019}. \\

@@

IThe final stage of our investigation deals with the plausible values of the dimensionless numbers $\mathcal{K}_c$ and $\mathcal{K}$, in comparison to the empirical measurements. We vary the body force coefficient ($k_n$) and the friction coefficient ($k_t$) to explore different values of $\mathcal{K}_c$ and $\mathcal{K}$, respectively.\\

We consider the empirical measurements from Ref.~\cite{helbing_2007} corresponding to the fundamental diagram obtained at the entrance of the Jamaraat bridge (see the inset in Fig.~\ref{flow_density_no_body_force}). Our aim is to reproduce the qualitative behavior of these measurements.\\

Figs. \ref{flow_density} show the pedestrian flow as a function of the global density (fundamental diagram). Fig.~\ref{flow_density_no_body_force} corresponds to $\mathcal{K}_c=$ 0 (this is \textit{i.e.} $k_n=0$) while Fig.~\ref{flow_density_body_force} corresponds to $\mathcal{K}_c=$68 corresponding to the original SFM with $k_n=1.2\,$E5. Each curve represents different friction values (see the caption for details).\\

According to the empirical measurements at Jamaraat, the flow slows down for high enough densities due to jamming. Notice that the original SFM (corresponding to $\mathcal{K}=$137) does not produce the expected slowing down for null body force ($\mathcal{K}_c=$0). However, when the body force is present (the original SFM) an ``U'' shape behavior occcurs for densities above 5 p$^{-2}$.\\

The increase in $\mathcal{K}$ to $\mathcal{K} = $ 685 (five times the original SFM value) slows down  the flux for densities above 5 p m$^{-2}$, regardless of the presence of the body force. Including the body force, however, produces a subtle increment in the flow for densities higher than 7 p m$^{-2}$ (see orange curve from Fig.~\ref{flow_density_no_body_force}). \\

A further increment of $\mathcal{K}$ to $\mathcal{K}=$137 (ten times the original SFM value) attains a plateau ($\rho>5$ p m$^{-2}$) before vanishing at very high densities. This occurs on either $\mathcal{K}_c=$ 0 and $\mathcal{K}_c = $ 68.\\

%It is worth mentioning that the SFM contemplates the possibility of overlapping between pedestrians (unlike the models stated in Refs.~\cite{kabalan_2017,jebrane_2019}). Despite this hard-to-interpret feature, the SFM can reproduce empirical behaviors of collective pedestrian dynamics.\\

These results suggest that although increasing $\mathcal{K}_c$ slows down the flux, it is still necessary to increase $\mathcal{K}$ (by increasing the friction coefficient $k_t$) to avoid an ``U'' shape behavior for extremely high densities. It is still a challenge to find the optimal dimensionless numbers  $(\mathcal{A},\mathcal{K},\mathcal{K}_c)$ for the fundamental diagram. Our inspection into the parameter set is a first approach to narrow down the search.\\

@@

We explored the effect on the pedestrians dynamics of the (sometimes neglected) body force in the framework of the SFM. We showed that the stiffness coefficient ($k_n$) has a significant impact on the evacuation dynamics (bottleneck) and also in the dynamics of pedestrians walking along a straight corridor.\\

In the bottleneck geometry, the evacuation time diminishes (pedestrians move faster) as pedestrians become stiffer along the explored desired velocities. This phenomenon occurs because stiffer pedestrians reduce the overlapping and hence the sliding friction intensity. This scenario releases more easily the pedestrians and reduces the probability of producing a cluster of pedestrians blocking the exit (blocking cluster). This leads to a more efficient evacuation dynamics. \\

The opposite behavior is obtained in the corridor geometry with respect to the bottleneck geometry. The major difference is that pedestrians are limited to the available space between walls in the corridor geometry.  Walls are not relevant in the bottleneck geometry. Thus, the overlap between pedestrians is controlled by the available space. But, stiffer pedestrians are more likely to get stuck. The whole crowd can be compared to a granular material. Granular materials can be disordered (amorphous) or ordered depending on how particles link to each other. In the present context, at low stiffness levels, the crowd appears disordered attaining a parabolic velocity profile.
If the stiffness level is high, the whole crowd appears ordered into a lattice (like a crystalline solid) with a uniform velocity profile that depends on the friction interaction with the walls.\\

Our efforts to ``tune'' the original SFM to empirical data (say, the fundamental diagram) moved us to explore the dimensionless parameter space.  We found that the qualitative behavior of empirical  data can be accomplished if the parameters are close to $\mathcal{K}_c < 68$, $\mathcal{K} = 685$ and $\mathcal{A}=14$. Nevertheless, this is a first attempt to arrive at suitable parameters, although we do not claim these to be optimal. We encourage further research to find dimensionless numbers that may better fit experimental data.\\
